Surrogate model for aligned-spin binary neutron star waveforms using gaussian process regression

### 2017-07-20 paper/paper.tex
Spin effects in tidal EOB

%% ________________________________________________________
\section{Aligned-spin, dynamical tides EOB model}
Ever since its inception~\cite{Buonanno:1998gg}, the effective-one-body (EOB) approach to the general-relativistic 2-body problem has proven successful in modeling the dynamics and GW emission of binaries of compact objects. State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match numerical-relativity (NR) simulations of spinning, nonprecessing binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to almost extremal for equal-mass) binaries. The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}. This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu}, which become especially relevant in the strong-field regime for 2-body systems that contain matter --- typically binary neutron stars (BNSs) or neutron-star/black hole binaries (NSBHs). The main effect of tides is to make the gravitational interaction more attractive with respect to the vacuum case. In particular, Refs.~\cite{Hinderer:2016eia,Steinhoff:2016rfi} built upon the aligned-spin BBH EOB model of Ref.~\cite{Taracchini:2013rva} and proposed a way to include the effect of dynamical tides. Neutron stars that are part of a compact-object binary will deform in the tidal field generated by the companion --- be it a black hole or another neutron star. Because of binary motion, the forcing tidal field varies at the orbital frequency. Thus, in the late stages of the inspiral, the characteristic $f$-mode frequency of the neutron star can be dynamically approached, resulting in a resonant excitation of such mode. The net effect is an amplification of tidal effects as compared to the adiabatic limit, which assumes that the $f$-mode frequency is much larger than the frequency of the forcing tidal field. Dynamical tidal effects are implemented in the EOB model through a modification of the radial potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime.

@@

As described above, an EOB waveform with dynamical tides depends on 11 intrinsic parameters: the mass ratio $0\leq q=M_B/M_A\leq1$, the dimensionless spin components along the angular momentum of the binary $-1 \leq \chi_{A,B} \leq1$, the adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, the $\ell=2,3$ $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$. We note that this list does not include the total mass $M$ of the system, since the point-mass part of the model is scale-invariant and tidal corrections only depend on $X_A=1/(1+q)$, $X_B=q/(1+q)$, and $\Lambda_2^{A,B}$, which is inherent to the PN analysis of tidal effects~\cite{Flanagan:2007ix}. To reduce the dimensionality of the intrinsic parameter space, we exploit universal (i.e., equation-of-state-independent) relations that are available in the literature. In particular, Eq.~(60) of Ref.~\cite{Yagi:2013sva} with the coefficients from the first row of Table I therein expresses $\Lambda_3^{A,B}$ as a function of $\Lambda_2^{A,B}$ with an accuracy to within about 10\%, depending on the equation of state and the value of $\Lambda_2^{A,B}$. Moreover, Eq. (3.5) of Ref.~\cite{Chan:2014kua} with the coefficients from the first column of Table I therein gives $\omega_{02}^{A,B}$ as a function of $\Lambda_2^{A,B}$ to within a few percent, while the same equation but with the coefficients from the second column of Table I therein gives $\omega_{03}^{A,B}$ as a function of $\Lambda_3^{A,B}$ to within a few percent. Thanks to these relations, the waveform only depends on 5 intrinsic parameters, which we denote as $\boldsymbol{x}=\{q, \chi_A,\chi_B, \Lambda_2^A,\Lambda_2^B\}$.

--Total mass is scaled out. This is fine, even up to zero amplitude point. Estimate the error in this approximation.

>>>>

%% ________________________________________________________
\section{Aligned-spin, dynamical tides EOB model}
Ever since its inception~\cite{Buonanno:1998gg}, the effective-one-body (EOB) approach to the general-relativistic 2-body problem has proven successful in modeling the dynamics and GW emission of binaries of compact objects. State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match hundreds of numerical-relativity (NR) simulations of spinning, nonprecessing binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to almost extremal for equal-mass) binaries. The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}. This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu}, which become especially relevant in the strong-field regime for 2-body systems that contain matter --- typically binary neutron stars (BNSs) or neutron-star/black hole binaries (NSBHs). The main effect of tides is to make the gravitational interaction more attractive with respect to the vacuum case. In particular, Refs.~\cite{Hinderer:2016eia,Steinhoff:2016rfi} built upon the aligned-spin BBH EOB model of Ref.~\cite{Taracchini:2013rva} and proposed a way to include the effect of dynamical tides. Neutron stars that are part of a compact-object binary will deform in the tidal field generated by the companion --- be it a black hole or another neutron star. Because of binary motion, the forcing tidal field varies at the orbital frequency. Thus, in the late stages of the inspiral, the characteristic $f$-mode frequency of the neutron star can be dynamically approached, resulting in a resonant excitation of such mode. The net effect is an amplification of tidal effects as compared to the adiabatic limit, which assumes that the $f$-mode frequency is much larger than the frequency of the forcing tidal field. Dynamical tidal effects are implemented in the EOB model through a modification of the radial potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime.

@@

We note that, to date, this is the only tidal EOB model that includes the effect of spin. More precisely, spin effects are those of the point-mass part of the model~\cite{Bohe:2016gbl}, that was calibrated to 141 NR simulations of aligned-spins BBHs. The model, however, does not account for the quadrupole that the neutron stars acquire when they are spinning. This is a genuine matter effect that appears at lower PN order as compared to the tidal effects that are currently modeled. The importance of this physical effect in the context of data analysis for ground-based GW interferometers is currently being studied~\cite{IanTanja}.

As described above, an EOB waveform with dynamical tides depends on 11 intrinsic parameters: the mass ratio $0\leq q=M_B/M_A\leq1$, the dimensionless spin components along the angular momentum of the binary $-1 \leq \chi_{A,B} \leq1$, the adiabatic quadrupolar tidal polarizabilities $\Lambda_2^{A,B}$, the adiabatic octupolar tidal polarizabilities $\Lambda_3^{A,B}$, the $\ell=2,3$ $f$-mode angular frequencies $\omega_{0\ell}^{A,B}$. To reduce the dimensionality of the intrinsic parameter space, we exploit universal (i.e., equation-of-state-independent) relations that are available in the literature. In particular, Eq.~(60) of Ref.~\cite{Yagi:2013sva} with the coefficients from the first row of Table I therein expresses $\Lambda_3^{A,B}$ as a function of $\Lambda_2^{A,B}$ with an accuracy to within about 10\%, depending on the equation of state and the value of $\Lambda_2^{A,B}$. Moreover, Eq. (3.5) of Ref.~\cite{Chan:2014kua} with the coefficients from the first column of Table I therein gives $\omega_{02}^{A,B}$ as a function of $\Lambda_2^{A,B}$ to within a few percent, while the same equation but with the coefficients from the second column of Table I therein gives $\omega_{03}^{A,B}$ as a function of $\Lambda_3^{A,B}$ to within a few percent. Thanks to these relations, the waveform only depends on 5 intrinsic parameters, which we denote as $\boldsymbol{x}=\{q, \chi_A,\chi_B, \Lambda_2^A,\Lambda_2^B\}$. We note that this list does not include the total mass $M$ of the system. Of course, the point-mass part of the model is scale-invariant. As to the inspiral-plunge phase, in this model tidal corrections only depend on $X_A=1/(1+q)$, $X_B=q/(1+q)$, and $\Lambda_2^{A,B}$, which is inherent to the post-Newtonian analysis of tidal effects~\cite{Flanagan:2007ix}. Finally, the simple model of the post-merger signal that we employ does not depend on $M$ either, although a more sophisticated description will have to include such dependence to account for the phenomenology that is observed in NR simulations of this phase.
